Article Title

### 2014-07-09 plos.tex
edits before submission

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser neuronal populations \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of  neural circuit organization.  However, estimating and interpreting large correlation matrices is statistically challenging.  Estimation can be improved by regularization, \ie by imposing a structure on the estimate.  The amount of improvement depends on how closely the assumed structure represents dependencies in the data. Therefore, the selection of the most efficient correlation matrix estimator for a given neural circuit must be determined empirically.  Importantly, the identity and structure of the most efficient estimator informs about the types of dominant dependencies governing the system.
We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix should be best modeled as the combination of a sparse graph of pairwise partial correlations representing local interactions and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate defined a graph of interactions. These interactions reflected the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this \sq{sparse + latent} estimator likely provides a more physiologically relevant representation of the functional connectivity in dense recordings than the sample correlation matrix.

@@

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix is defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; the $p\times 1$ vector $x$ is a single observation of the firing rates of $p$ neurons over time $\Delta t$; and $\mu$ is the vector of expected firing rates.

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is a $p\times 1$ vector of firing rates in time bin $t$, and an independent unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix},
\begin{equation}\label{eq:sample}
	C_{\sf sample} = \frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
\end{equation}
where $n$ is the size of $T$, yields an unbiased estimate of the covariance matrix so that $\E{C_{\sf sample}}=\Sigma$.

When the mean is estimated from the same sample, $C_{\sf sample}$ is biased toward zero.  However, in all the cases when the unbiasedness is required in our study, the mean will be estimated from a separate sample.

@@

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:1990}. Partial correlations figure prominently in probablistic \emph{graphical modelling} wherein the joint distribution is explained by sets of two-way interactions \cite{Whittaker:1990}. For the multivariate Gaussian distribution, zero partial correlations indicate conditional independence of the pair, implying a lack of direct interaction \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1990,Baba:2004}. As neural recordings become increasingly dense, partial correlations may prove useful as indicators of conditional independence (lack of functinonal connectivity) between pairs of neurons.

@@

As the size of recorded neuronal populations increases, without substantial increases in recording durations, the \emph{condition number} of the sample covariance matrix also increases, making the partial correlation estimates \emph{ill-conditioned}: small errors in covariance estimates translate into much larger errors in partial correlations. With massively multineuronal recordings, partial correlations cannot be estimated without \emph{regularization} \cite{Ledoit:2004,Schafer:2005}.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure01.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
    {\bf Row 1.} Graphical representations of the target estimates of the four respective regularized covariance matrix estimators.  Recorded neurons are represented by green spheres and latent units by light-shaded spheres.  Edges represent non-zero partial correlations, \ie \sq{interactions}.
        {\bf Row 1, A}.  For estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear dependencies. 
        {\bf  Row 1, B.} For estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), representing systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. For estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from a sparse set of  linear associations between the observed units. 
        {\bf  Row 1, D}.  For estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to sparse associations between the recorded cells \emph{and} with several latent units. \\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with four latent units, {\bf C.}  a correlation matrix with 67\%  off-diagonal zeros in its inverse, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-3 (\ie three latent units) matrix and a sparse matrix with 76\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=500$ drawn from simulated random processes with respective correlation matrices shown in Row 2.  The structure of the sample correlation matrix is difficult to discern.
\\
{\bf Row 4:} Correlation matrix estimates computed by estimators with matching structure from the same data as the sample correlation matrices in Row 3. The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:loss}) for the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with structure that matches the true model converged to zero faster than the other estimators. Some exceptions occur when the sample is too small to reveal the structure in the data. In these cases estimators with simpler structures can be more efficient than estimators with true structure.
\\
{\bf Row 6:} Validation losses for the five estimators relative to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:1}
\end{FPfigure}

The target estimate of estimator $C_{\sf diag}$ is the diagonal matrix $D$ containing estimates of neurons' variances. Regularization is achieved by linear \emph{shrinkage} of the unbiased estimate $C_{\sf sample}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}\label{eq:c-diag}
C_{\sf diag} = (1-\lambda) C_{\sf sample} + \lambda D
\end{equation}
The structure imposed by $C_{\sf diag}$ favors (performs better  on) populations   with no linear associations between the neurons (\figref{1}{Row 1, A}).  If a predominant majority of sample correlations are spurious, $C_{\sf diag}$ is expected to be more efficient than other estimators.

Estimator $C_{\sf factor}$ approximates the covariance matrix by the factor model $L + D$, where $L$ is the $p\times p$ positive semidefinite matrix with low-rank and $D$ is a diagonal matrix of independent variances. The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = L + (1-\lambda)D + \lambda \bar D,
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

Although the two functions are similar in form, they differ conceptually: The log-likelihood $L(\Sigma|C_{\sf sample})$ is a function of the parameter $\Sigma$ given the sample covariance matrix $C_{\sf sample}$ whereas the loss function $\loss{C,\Sigma}$ expresses the deviation of an arbitrary estimate $C$ from the truth $\Sigma$. Minimum expected loss may be attaned at a point far removed from the point of maximum likelihood point given the same sample even when the normal distribution can be assumed.

@@

The normal loss (Eq.~\ref{eq:loss}) is particularly suitable because it is additive in its second argument, \ie
\begin{equation*}\label{eq:additivity}
    \loss{C,z_1} + \loss{C,z_2} \equiv \loss{C,z_1+z_2}
\end{equation*}
Then, once $C$ is fixed, the validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E{C_{\sf sample}^\prime}}=\loss{C,\Sigma}.
\end{equation*}

The property of additivity does not hold for other popular loss functions such as Stein's entropy loss or various quadratic losses; their validation losses are biased estimates of the loss function. For these loss functions, cross-validation does not generally evaluate the discrepancy from truth.

The loss function above assumes a single covariance matrix across all conditions in the dataset.  However, neuroscientists often estimate a common correlation matrix across multiple stimulus conditions when the means and variances of responses are conditioned on the stimulus while correlations are assumed constant across all conditions \cite{Vogels:1989, Ponce:2013}. In this case, the stimulus-dependent means and variances are estimated from the training dataset along with a common correlation matrix. The adaptation of the normal loss function to this case is described in Methods.

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the testing sample with the other $K-1$ serving as the training sample. The validation losses from all such \sq{folds} are averaged to produce the \emph{cross-validation loss}.

Cross-validation loss accurately reproduced the relationships between the excess losses of the estimators (\figref{1}{Row 6}).

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C}). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{D}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low. The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean of 0.018 (\figref{2}{F} and \figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals of seven cells, downsampled to 20 Hz, out of the 292 total recorded cells. Spiking activity inferred by non-negative deconvolution is shown by red ticks below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 292 cells from the imaged site. Only spiking cells are shown. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients. The red line indicates the mean correlation coefficient of 0.018.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

@@

\begin{figure}[!ht]
    \begin{center}
    \includegraphics{./figures/src/Fig3.eps}
    \end{center}
    \caption{
   {\bf Performance of estimator $C_{\sf sparse+latent}$ with respect to the normal loss function (eq.~\ref{eq:loss}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} 
    Covariance estimators $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ produced consistently greater validation losses than $C_{\sf sparse+latent}$ ($p<10^{-2}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). The box plots indicate the $25^{th}$, $50^{th}$, and $75^{th}$ percentiles with the whiskers extending to the minimum and maximum values after excluding the outliers marked with `+'. 
} \label{fig:3}
\end{figure}

@@

\paragraph{Structure of $C_{\sf sparse+latent}$ estimates}
The estimator $C_{\sf sparse+latent}$ separates two sources of correlations: a network of linear interactions between pairs of neurons and a set of latent units reflecting common fluctuations across the entire population.

We examined the composition of the $C_{\sf sparse+latent}$ estimates at each imaged site (Fig.~\ref{fig:4} and Fig.~\ref{fig:5}). Although the regularized estimates were similar to the sample correlation matrix (\figref{4}{A and B}), the corresponding partial correlation matrices differed substantially (\figref{4}{C and D}). The partial correlation matrix of the regularized estimate was the sum of a sparse component and a low-rank component (\figref{4}{E}).  With the sparse+latent estimator, we distinguish the pairwise partial correlations computed according to Eq.~\ref{eq:partial}, which include both the sparse and the latent components, from the \emph{sparse partial correlations}, which are computed from the sparse component (Eq.~\ref{eq:c-sl}) alone and represent the local, direct interactions. Thus panels \figref{4}{C and D} show partial correlations whereas panel \figref{4}{E and F} depict the sparse partial correlations.

@@

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 27 sites were about 5 times lower than the average sample correlations (\figref{5}{C}). This suggests that correlations between neurons build up from large numbers of small interactions with other neurons. Furthermore, the average partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.45 whereas that of the average partial correlations was 0.29, with larger populations exhibiting more similar values of the average partial correlations than the smaller populations.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics[width=17.35cm]{./figures/Figure04.pdf}
    \end{center}
    \caption{{\bf Example of the structure revealed by $C_{\sf sparse+latent}$ the sparse+latent estimator.}
    {\bf A and B.} The regularized estimate $C_{\sf sparse+latent}$ closely approximates the sample correlation matrix $C_{\sf sample}$. 
    {\bf C and D.} However, the partial correlation matrices from the two estimates differ substantially.
    {\bf E.} The partial correlation matrix of the regularized estimate is decomposed into a sparse component with 92.8\% off-diagonal zeros (bottom-left) and low-rank component of rank 72 (top-right).
    {\bf F.} The sparse component of the regularized partial correlation matrix had little resemblance to the sample correlations: the gray interval indicates the range of correlations containing 92.8\% of cells pairs, equal to the fraction of zeros in the sparse partial correlation matrix. The significant correlations were outside this interval. Yet 72.3\% of the interactions inferred by $C_{\sf sparse+latent}$  linked pairs of neurons whose correlation was below the threshold. Only the remaining 27.7\% of the interactions overlapped with sample correlations above the threshold.
    {\bf G.} A graphical depiction of the positive (green) and negative (magenta) partial correlations as edges between observed neurons. The line density is proportional to the magnitude of the correlation.
    {\bf H.} A subset of neurons from the center of the cluster shown in {\bf G} showing the regularized partial correlations.
    {\bf I.} The same subset with sample correlations thresholded to match the sparsity of the regularized interactions.
}
\label{fig:4}
\end{FPfigure}

Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013, Sadovsky:2014}.  We therefore compared the interactions in the sparse component of $C_{\sf sparse+latent}$ to those obtained from the sample correlations thresholded to match the connectivity of the sparse component of $C_{\sf sparse+latent}$.  The networks revealed by the two methods differed substantially. In the example site with 7.2\% connectivity in $C_{\sf sparse+latent}$, only 27.7\% of the connections coincided with the above-threshold sample correlations (\figref{4}{F, H, and I}). In particular, most of the inferred negative interactions corresponded to low sample correlations (\figref{4}{F}) where high correlations were expected given the rest of the correlation matrix.

@@

\paragraph{Data processing}
All data were processed in MATLAB using the DataJoint data processing chain toolbox \cite{DataJoint:2014}.

@@

\paragraph{Cross-validation}
To compare the performance of the estimators, we used conventional 10-fold cross validation: Trials were randomly divided into 10 subsets with approximately equal numbers of trials of each condition in each subset. Each subset was then used as the testing sample with the rest of the data used as the training sample for estimating the covariance matrix. The resulting validation losses were averaged to produce the \emph{validation loss} reported throughout the paper.

>>>>

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser neuronal populations \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of  neural circuit organization.  However, estimating and interpreting large correlation matrices is statistically challenging.  Estimation can be improved by regularization, \ie by imposing a structure on the estimate.  The amount of improvement depends on how closely the assumed structure represents dependencies in the data. Therefore, the selection of the most efficient correlation matrix estimator for a given neural circuit must be determined empirically.  Importantly, the identity and structure of the most efficient estimator informs about the types of dominant dependencies governing the system.
We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these densely sampled recordings, the correlation matrix should be best modeled as the combination of a sparse graph of pairwise partial correlations representing interactions between the observed neuronal pairs and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate defined a graph of interactions. These interactions reflected the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this \sq{sparse + latent} estimator likely provides a more physiologically relevant representation of the functional connectivity in densely sampled recordings than the sample correlation matrix.

@@

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix is defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; the $p\times 1$ vector $x$ is a single observation of the firing rates of $p$ neurons in a time bin of some duration; and $\mu$ is the vector of expected firing rates.

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is a $p\times 1$ vector of firing rates in time bin $t$, and an independent unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix},
\begin{equation}\label{eq:sample}
	C_{\sf sample} = \frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
\end{equation}
where $n$ is the number of time bins in $T$, yields an unbiased estimate of the covariance matrix so that $\E{C_{\sf sample}}=\Sigma$.

When the mean is estimated from the same sample, $C_{\sf sample}$ becomes biased toward zero.  However, in all the cases when the unbiasedness is required in our study, the mean will be estimated from a separate sample.

@@

The \emph{partial correlation} between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:1990}. Partial correlations figure prominently in probablistic \emph{graphical modelling} wherein the joint distribution is explained by sets of two-way interactions \cite{Whittaker:1990}. For the multivariate Gaussian distribution, zero partial correlations indicate conditional independence of the pair, implying a lack of direct interaction \cite{Dempster:1972, Whittaker:1990}. More generally, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1990,Baba:2004}. As neural recordings become increasingly dense, partial correlations may prove useful as indicators of conditional independence (lack of functinonal connectivity) between pairs of neurons.

@@

As the size of recorded neuronal populations increases, without substantial increases in recording durations, the \emph{condition number} of the sample covariance matrix also increases, making the partial correlation estimates \emph{ill-conditioned}: small errors in covariance estimates translate into much greater errors in partial correlations after the inversion. With massively multineuronal recordings, partial correlations cannot be estimated without \emph{regularization} \cite{Ledoit:2004,Schafer:2005}.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure01.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
    {\bf Row 1.} Graphical models of the target estimates of the four respective regularized covariance matrix estimators.  Recorded neurons are represented by the green spheres and latent units by the lightly shaded spheres.  Edges represent non-zero partial correlations, \ie \sq{interactions}.
        {\bf Row 1, A}.  For estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear dependencies. 
        {\bf  Row 1, B.} For estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), representing systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. For estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from a sparse set of  linear associations between the observed units. 
        {\bf  Row 1, D}.  For estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to sparse associations between the recorded cells \emph{and} due to several latent units. 
	\\
	{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with four latent units, {\bf C.}  a correlation matrix with 67\%  off-diagonal zeros in its inverse, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-3 matrix (\ie three latent units) and a sparse matrix with 76\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=500$ drawn from simulated random processes with respective correlation matrices shown in Row 2.  The structure of the sample correlation matrix is difficult to discern by eye.
\\
{\bf Row 4:} Estimates computed from the same data as in Row 3 using structured estimators of the correct type, optimized by cross-validation.  The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:loss}) for the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with structure that matches the true model converged to zero faster than the other estimators. 
\\
{\bf Row 6:} Validation losses for the five estimators relative to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.  Differences in validation loss approximate differences in true loss.
    }
    \label{fig:1}
\end{FPfigure}

The target estimate of estimator $C_{\sf diag}$ is the diagonal matrix $D$ containing estimates of neurons' variances. Regularization is achieved by linear \emph{shrinkage} of the unbiased estimate $C_{\sf sample}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}\label{eq:c-diag}
C_{\sf diag} = (1-\lambda) C_{\sf sample} + \lambda D
\end{equation}
The structure imposed by $C_{\sf diag}$ favors (performs better  on) populations   with no linear associations between the neurons (\figref{1}{Row 1, A}).  If sample correlations are largely spurious, $C_{\sf diag}$ is expected to be more efficient than other estimators.

Estimator $C_{\sf factor}$ approximates the covariance matrix by the factor model $L + D$, where $L$ is the $p\times p$ positive semidefinite matrix of a low rank and $D$ is a diagonal matrix of independent variances. This approximation is the basis for \emph{factor analysis} \cite{Anderson:2003}. The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = L + (1-\lambda)D + \lambda \bar D,
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$ to shrink the independent variances toward their mean value $\bar D$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

Although the two functions are similar in form, they differ conceptually: The log-likelihood $L(\Sigma|C_{\sf sample})$ is a function of the parameter $\Sigma$ given the sample covariance matrix $C_{\sf sample}$ whereas the loss function $\loss{C,\Sigma}$ expresses the deviation of an arbitrary estimate $C$ from the truth $\Sigma$. Minimum expected loss may be found at a point that is far removed from the point of maximum likelihood given the same data.

@@

The normal loss (Eq.~\ref{eq:loss}) is particularly suitable because it is additive in its second argument, \ie
\begin{equation*}\label{eq:additivity}
    \loss{C,Z_1} + \loss{C,Z_2} \equiv \loss{C,Z_1+Z_2}
\end{equation*}
Then, with fixed $C$, validation loss is an unbiased estimate of true loss:
\begin{equation*}
    \E{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E{C_{\sf sample}^\prime}}=\loss{C,\Sigma}.
\end{equation*}

The property of additivity does not hold for other popular loss functions such as Stein's entropy loss or various quadratic losses; their validation losses are biased estimates of the loss function and validation does not exactly evaluate a discrepancy from truth.

The loss function above assumes a single covariance matrix across all conditions in the dataset.  However, neuroscientists often estimate a common correlation matrix across multiple stimulus conditions when the variances of responses are conditioned on the stimulus \cite{Vogels:1989, Ponce:2013}. In this case, the stimulus-dependent means and variances are estimated from the training dataset along with a common correlation matrix. An adaptation of the normal loss function to this case is described in Methods.

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the testing sample with the other $K-1$ subsets serving as the training sample. The validation losses from all such \sq{folds} are averaged to produce the \emph{cross-validation loss}.

Cross-validation loss accurately reproduced the relationships between the excess losses of the estimators (\figref{1}{Row 6}), confirming that cross-validation can evaluate true loss.

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C and D}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C}). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{D}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the correlation matrix.   The sample correlation coefficients were largely positive and low (\figref{2}{E and F}). The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean across sites of 0.018 (\figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals of seven cells, downsampled to 20 Hz, out of the 292 total recorded cells. Spiking activity inferred by non-negative deconvolution is shown by red ticks below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 292 cells from the imaged site. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients in one site. The red line indicates the mean correlation coefficient of 0.020.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

@@

\begin{figure}[!ht]
    \begin{center}
    \includegraphics{./figures/src/Fig3.eps}
    \end{center}
    \caption{
   {\bf Performance of estimator $C_{\sf sparse+latent}$ with respect to the normal loss function (eq.~\ref{eq:loss}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} 
    Covariance estimators $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ produced consistently greater validation losses than $C_{\sf sparse+latent}$ ($p<0.01$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). The box plots indicate the $25^{th}$, $50^{th}$, and $75^{th}$ percentiles with the whiskers extending to the minimum and maximum values after excluding the outliers marked with `+'. 
} \label{fig:3}
\end{figure}

@@

\paragraph{Structure of $C_{\sf sparse+latent}$ estimates}
We examined the composition of the $C_{\sf sparse+latent}$ estimates at each imaged site (Fig.~\ref{fig:4} and Fig.~\ref{fig:5}). Although the regularized estimates were similar to the sample correlation matrix (\figref{4}{A and B}), the corresponding partial correlation matrices differed substantially (\figref{4}{C and D}). The estimates separated two sources of correlations: a network of linear interactions expressed by the sparse component of the inverse and latent units expressed by the low-rank components of the inverse (\figref{4}E). The sparse partial correlations revealed a network that differed substantially from the network composed of the greatest coefficients in the sample correlation matrix (\figref{4}{F, G, H, and I}).

@@

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 27 sites were about 5 times lower than the average sample correlations (\figref{5}{C}). This suggests that correlations between neurons build up from multiple chains of smaller interactions. Furthermore, the average partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.45 whereas that of the average partial correlations was 0.29, with larger populations exhibiting more similar values of the average partial correlations than the smaller populations.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics[width=17.35cm]{./figures/Figure04.pdf}
    \end{center}
    \caption{{\bf Structure revealed by $C_{\sf sparse+latent}$.}
    {\bf A, B.} The regularized estimate $C_{\sf sparse+latent}$ closely approximates the sample correlation matrix $C_{\sf sample}$. 
    {\bf C, D.} The partial correlation matrices from the two estimates differ substantially.
    {\bf E.} The partial correlation matrix of the regularized estimate is decomposed into a sparse component with 92.8\% off-diagonal zeros (bottom-left) and low-rank component of rank 72 (top-right).
    {\bf F.} The sparse component of the regularized partial correlation matrix had little resemblance to the sample correlations: The gray region indicates the range of correlations containing 92.8\% of cells pairs, equal to the fraction of zeros in the sparse partial correlation matrix. Correlation coefficients outside this interval formed the network of greatest correlations.  This network differed from the sparse component of the $C_{\sf sparse+latent}$:  Only 27.7\% of the highest correlations coefficients outside the gray regions coincided with interactions inferred by $C_{\sf sparse+latent}$.
    {\bf G.} A graphical depiction of the positive (green) and negative (magenta) sparse partial correlations as edges between observed neurons. The line density is proportional to the magnitude of the partial correlation.
    {\bf H.} A subset of neurons from the center of the cluster shown in {\bf G} showing the sparse partial correlations.
    {\bf I.} The same subset of neurons with edges indicating sample correlations thresholded to match the sparsity of the sparse partial correlation. These edges correspond to the sample correlation coefficients outside the gray region in panel F.
}
\label{fig:4}
\end{FPfigure}

Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013, Sadovsky:2014}.  We therefore compared the interactions in the sparse component of $C_{\sf sparse+latent}$ to those obtained from the sample correlations thresholded to the same level of connectivity. The networks revealed by the two methods differed substantially. In the example site with 7.2\% connectivity in $C_{\sf sparse+latent}$, only 27.7\% of the connections coincided with the above-threshold sample correlations (\figref{4}{F, H, and I}). In particular, most of the inferred negative interactions corresponded to low sample correlations (\figref{4}{F}) where high correlations should be expected given the rest of the correlation matrix.

@@

\paragraph{Data processing}
All data were processed in MATLAB using the DataJoint data processing chain toolbox (http://datajoint.github.com).

@@

\paragraph{Cross-validation}
To compare the performance of the estimators, we used conventional 10-fold cross-validation: Trials were randomly divided into 10 subsets with approximately equal numbers of trials of each condition in each subset. Each subset was then used as the testing sample with the rest of the data used as the training sample for estimating the covariance matrix. The resulting validation losses were averaged to produce the \emph{validation loss} reported throughout the paper.
